Liveness-Based Garbage Collection for Lazy Languages

### 2015-11-26 paper/paper.tex
Updated liveness eqns
latest on 26th

\documentclass[9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
%\newcommand{\mycomment}[1]{}
\definecolor{Myblue}{rgb}{.2,0,1}
%\newcommand{\comment}[1]{{\color{Myblue}{(#1)}}}
%\newcommand{\cblue}[1]{{\color{blue}{#1}}}
%\newcommand{\blankout}[1]{}

%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{\mbox{S}}   
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}
% --- Author Metadata here ---
%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
%over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
\cmt{{
\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
section.
%
\author{
% 1st. author
\alignauthor K. Prasanna Kumar \\
       \affaddr{IIT Bombay,}\\
       \affaddr{Mumbai 400076, India}
       \email{prasannak@cse.iitb.ac.in}
% 2nd. author
\alignauthor Amey Karkare\\
       \affaddr{IIT Kanpur}\\
       \affaddr{Kanpur 208016, India}
       \email{karkare@cse.iitk.ac.in}
% 3rd. author
\alignauthor Amitabha Sanyal \\
       \affaddr{IIT Bombay}\\
       \affaddr{Mumbai 400076, India}\\
       \email{as@cse.iitb.ac.in}
}
}}
\authorinfo{Double Blind Review}{}{}
\maketitle

\begin{abstract}
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs.  To do this, we  analyze programs for
liveness of  heap-allocated data.  The accompanying  garbage collector
is modified  to use the result  of the analysis to  preserve only live
data---a subset  of reachable data---during garbage  collection.  This
results  in an  increase  in the  garbage  reclaimed and  consequently
reduces the  memory footprint of  programs.  While this  technique has
already been shown to yield  benefits for eager first-order languages,
the lack of a statically determinable execution order and the presence
of closures at runtime pose additional challenges for lazy languages.
This requires changes  both in the earlier liveness  analysis and also
in the  garbage collection scheme  required to reclaim memory  that is
not live.

The implementation  of our  analysis annotates each  potential garbage
collection  point  in   the  program  with  a   set  of  deterministic
finite-state automata (DFA) describing the  liveness at that point. We
also implement a copying collector that  uses the DFA to preserve live
objects.   Our  experiments  confirm  that  liveness-based  collection
always increases  the garbage  reclaimed, consequently  decreasing the
number of collections.  In addition,  it reduces the execution time of
several programs.  Experiments  also reveal that the  best results are
obtained  for   a  mixed  garbage   collection  scheme  that   does  a
liveness-based    collection   for    fully    evaluated   data    and
reachability-based   collection  for   closures.   On our benchmark
programs, we obtain a decrease of up to 45\% in the number of
collections, and a decrease of up to 88\% in the number of dragged
cells.  Even though liveness based collector uses an expensive
traversal scheme, the execution time is competitive (0.2X - 2.1X) when
compared to a reachability based collector.
\end{abstract}

@@

Functional programs use dynamically allocated memory extensively.  The
allocation  is either  explicit (while using  constructors, for example) or
implicit (as while creating  closures).  Programs  written in  lazy
functional languages put additional  demands on memory as they require
closures to  be carried  from the  point of creation  to the  point of
evaluation.

@@

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The analysis  has two  parts.  First,  each function  is
summarized in a context-sensitive manner as a {\em demand transformer}
that transforms a {\em symbolic} demand  on its body to demands on its
arguments.  This summary  is then used to step  through function calls
during analysis.   Second, the concrete  demand on a function  body is
obtained through a 0-CFA-like conservative approximation that combines
the  demands on  all the  calls to  the function.   The result  of the
analysis  is an  annotation of  each program  point with  finite-state
automata capturing the liveness of  variables at that point. Depending
on the program point where  garbage collection has been triggered, the
GC consults a set of  automata to restrict reachability during marking.
This results in an increase  in the garbage reclaimed and consequently
in fewer garbage collections.

While this idea has been shown to be effective~\cite{asati14lgc} for a
first-order {\em  eager} language, a straightforward  extension of the
earlier  technique  is  not  possible  for  lazy  languages,  where  a
heap-allocated   object   may   be  an   unevaluated   expression   or
closure. Since  data is made  live by  evaluation of closures,  and in
lazy languages  the place in  the program where this  evaluation takes
place cannot  be statically determined, laziness  complicates liveness
analysis itself.   In addition, apart  from evaluated data,  we extend
liveness-based collection  to closures---closures  with no  future use
should also  be collected.  This  involves tracking closures  to their
points of  creation at  runtime from where  we obtain  their liveness.
This requires significant modification to the earlier GC scheme.

Experiments    with   a    single    generation   copying    collector
(Section~\ref{sec:experiments})   confirm  the   expected  performance
benefits.  Liveness-based collection results in an increase in garbage
reclaimed.  As  a consequence, there  is a  decrease in the  number of
collections (up to 45\%) and a decrease  (up to 88\%) in the number of
{\em  dragged}  cells  (cells  that  survive a  GC  after  their  last
use). Besides, there is a reduction  in the overall execution time for
some programs.  Our  experiments show that a mixed  strategy, where we
do liveness based GC up to the closure boundary and reachability based
GC for the arguments (free variables) of the closure, gives us similar
benefits as  a fully  liveness based  GC (where  the arguments  of the
closure  are  also  collected  using liveness)  with  lesser  overhead
(Section~\ref{sec:strategies}).

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\begin{figure}[t!]
  \psset{unit=1mm}
  \begin{pspicture}(90,52)(0,-52)
    %\psframe(90,48)(0,-48)
    \renewcommand{\arraystretch}{1}
    \begin{tabular}{@{}c@{}}
      %\psframebox
          {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
            \UFL\
            \UNL{0} (\DEFINE\ (\length\  \pl)
	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
            (\PRIM\ 1\ (\length (\CDR\  \pl)))))
            \UNL{0}
	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
	    \listb
	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\  \lista) (\append\
            (\CDR\  \lista)
            \listb))))
            \UNL{0}
            \UNL{0} (\LET\ \px\
            $\leftarrow$\ (\CONS\ $5$
            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	    \UNL{2}
            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
            (\length\ \pz))))))
	  \end{uprogram}
      }}
      \\ \\
      (a) Example program. \\  \\
%      \includegraphics[width=.45\textwidth]{motiv-example}
      \input{mem-graph} \\ \\
      \renewcommand{\arraystretch}{.9}
      \begin{tabular}[t]{p{0.9\columnwidth}}
        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
        denotes a closure. Thick edges denote live links. Traversal
        stops at edges marked $\times$ during garbage collection for
        an ideal collector.
      \end{tabular}
    \end{tabular} 
  \end{pspicture}
%  \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

Figure~\ref{fig:mot-example} shows an example program and the state of
the heap at  the program point $\pi$, i.e. just  before the evaluation
of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
node either  represents atomic  values ($\NIL$,  integers etc.),  or a
\CONS\ cell containing $\CAR$ and $\CDR$  fields, or a chunk of memory
representing  an  unevaluated  value   also  called  a  {\em  closure}
(represented  by  hatched  boxes).   Edges   in  the  graph  are  {\em
  references} and emanate from variables  or fields.  The figure shows
the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
triggered    by   the    test   (\NULLQ~(\CAR~\pz))    in   the
\SIF\ expression.

The edges  shown by thick  arrows are those  which are live  at $\pi$.
Ideally, a cell should be  preserved during garbage collection only if
it is reachable from  the root set through a path  of live edges.  All
other  cells can  be reclaimed.   Thus if  a garbage  collection takes
place at $\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b),
an ideal liveness-based  collector (LGC) will preserve  only the cells
referenced by  $\pz$, $(\CDR~ \pz)$  and (the cells  constituting) the
closure   referenced   by   $(\CDR~(\CDR~\pz))$.    In   contrast,   a
reachability-based collector (RGC) would have preserved all cells.

While an ideal garbage collector is  impractical, in this work we show
that  static analysis  of  heap data  can help  practical garbage
collectors in reclaiming more  garbage.  The specific contributions of
this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
  functional language  and prove its correctness.  We show  how to
  compute  function  summaries  in  a context  independent  manner  to
  achieve  a  context  sensitive  analysis. We  further  describe  the
  transformation of the  liveness equations to grammar rules  in a way
  that the language  accepted by the resulting  grammar describes live
  heap locations.
\item We prove  the undecidability of the grammar  that results from
  the  analysis.   To overcome  this,  we  present an  algorithm  to
  transform grammars  to finite  state automata  (DFA) such  that the
  language accepted by the  DFA over-approximates the liveness (and
  hence a safe approximation).
\item We have implemented a garbage  collector that can use the DFA
  to  retain live  cells that  also include  closures. We  have done
  extensive  experiments with  this garbage  collector and  obtained
  empirical results  that show  the effectiveness  of liveness-based
  garbage collection.
\end{itemize}

@@

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language  considered  in our  analysis  and  a small-step  operational
semantics  for it.   We present  our liveness  analysis for  this lazy
language  in Section~\ref{sec:liveness}.   Section~\ref{sec:computing}
describes  the  formulation of  the  liveness  as  a grammar  and  its
approximation  to   DFA.  It  also   contains  a  proof   showing  the
undecidability  of  the  precise  liveness in  our  formulation  which
necessitates  the approximation  to DFA.   Section~\ref{sec:GC-scheme}
discusses different ways of handling closures and their pros and cons.
We  report our  experimental results  in Section~\ref{sec:experiments}
along with some observations.  Section~\ref{sec:relatedwork} discusses
previous  work   related  to  garbage  collection   and  liveness  and
Section~\ref{sec:conclusion}   discusses   possible   extensions   and
concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  $\LET$s  in our
language  are lazy;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ may occur in  $s$. This enables creation of graph-like
structures in  a pure language. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  the  reasoning   about  the  program.   The
$\SRETURN$  and   \SIF\  expressions  and  strict   operators  trigger
evaluation of closures.

The body  of a function ${\mathit  f}$ is denoted  as $e_{\mathit f}$.
We   shall   extend   this   convention   to   the   main   expression
$e_\mainpgm$. We assume that each program has a distinguished function
\mainpgm\      with      the      definition     $(\DEFINE\      ({\tt
  \mainpgm})\  e_\mainpgm)$ and  the execution  of the  program starts
with the call to \mainpgm.  We write $\pi\!:\!e$ to associate the label
$\pi$ (not  part of the language  syntax) with the  program point just
before expression $e$.

>>>>

\documentclass[preprint,9pt]{sigplanconf}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{lineno}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\usepackage{subfigure}
\usepackage[boxed]{algorithm2e}
\usepackage{makecell}
\usepackage{balance}

\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it
      %probably doesn't scale anyway.
%\newcommand{\mycomment}[1]{} 
\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}

\newcommand{\comment}[1]{{\color{Myblue}{#1}}}
\newcommand{\warning}[1]{{\color{Myred}{#1}}}
%\newcommand{\blankout}[1]{}

\newcommand{\cred}[1]{\psframebox[linestyle=none, fillcolor=lightgray,fillstyle=solid,framesep=0.5pt]{#1}}
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } 
\newcommand{\stk}{\mbox{S}}   
\newcommand{\ID}{\mbox{$\mathbf{ id}$}}
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}
\linenumbers
% --- Author Metadata here ---
%\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be
%over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data
%(0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
%\cmt{{
%
%\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
%section.
%
%\author{
% 1st. author
\authorinfo {Prasanna Kumar. K} 
       {IIT Bombay, Mumbai 400076, India}
       {prasannak@cse.iitb.ac.in}
% 2nd. author
\authorinfo {Amitabha Sanyal} 
       {IIT Bombay, Mumbai 400076, India}
       {as@cse.iitb.ac.in}
% 3rd. author
\authorinfo {Amey Karkare}
       {IIT Kanpur, Kanpur 208016, India}
       {karkare@cse.iitk.ac.in}
% 3rd. author
%% \alignauthor Amey Karkare\\
%%        \affaddr{IIT Kanpur}\\
%%        \affaddr{Kanpur 208016, India}
%%        \email{karkare@cse.iitk.ac.in}
%}
%}}
%\authorinfo{Double Blind Review}{}{}
\maketitle

\begin{abstract} 
We consider  the problem of reducing  the memory required to  run lazy
first-order functional programs. We  address the problem by analyzing
programs  for liveness  of  heap-allocated data.   The  result of  the
analysis is  used to preserve  only live data---a subset  of reachable
data---during garbage  collection.  The result  is an increase  in the
garbage reclaimed and  a  reduction in  the peak memory requirement 
of programs.   While this  technique has already  been shown  to yield
benefits for  eager first-order  languages, the  lack of  a statically
determinable execution  order and  the presence  of closures  pose new
challenges  for lazy  languages.  These  requires changes  both in  the
liveness analysis itself and in the design of the garbage collector.

%% We use the  result of the analysis to annotate  each potential garbage
%% collection  point  in   the  program  with  a   set  of  deterministic
%% finite-state automata (DFA) describing the liveness at that point.
We
implement a copying collector that uses the results of the liveness
analysis to preserve live
objects, both evaluated (i.e.  in WHNF) and closures.  Our experiments
confirm  that  for  programs  running with  a  liveness-based  garbage
collector,   there  is   a   significant  decrease   in  peak   memory
requirements.   In  addition,   a  sizable  reduction  in   the  number  of
collections  ensures that  in spite  of using  a more  complex garbage
collector,  the execution  time  of  a program  remains comparable  with
the same program running with a reachabilty-based collector.
\warning{On our benchmark programs, we obtain a decrease of up to 75\%
  in the number of collections, and up to 50\% decrease in peak memory
  requirements.}
\end{abstract}

@@

Functional programs make extensive use of dynamically allocated
memory.  The
allocation  is either  explicit (using  constructors, for example) or
implicit (creating  closures).  Programs  written in  lazy
functional languages put additional  demands on memory as they require
closures to  be carried  from the  point of creation  to the  point of
evaluation.

@@

In this paper,  we propose the use of liveness  analysis of heap cells
for garbage collection in a lazy first-order functional language.  The
central notion in our analysis  is a generalization of liveness called
{\em  demand}---the  pattern  of  future  uses  of  the  value  of  an
expression.   The  analysis has  two  parts.   First, we  calculate  a
context-sensitive  summary   of  each   function  as  a   {\em  demand
  transformer} that transforms a {\em  symbolic} demand on its body to
demands on its  arguments.  This summary is used  to step through
function  calls during  analysis.  Second,  the concrete  demand on  a
function   body  is   obtained  through   a  0-CFA-like   conservative
approximation  that combines  the  demands  on all  the  calls to  the
function.   The result  of the  analysis is  an annotation  of certain
program points  with deterministic finite-state  automata (DFA)  capturing the  liveness of
variables  at  these points.  Depending  on  the program  point  where
garbage  collection has  been  triggered,  the garbage collector consults  a set  of
automata to restrict reachability during  marking.  This results in an
increase in  the garbage reclaimed  and consequently in  fewer garbage
collections.

While this idea has been shown to be effective for a
first-order {\em  eager} language~\cite{asati14lgc}, a straightforward  extension of the
technique   is  not   possible  for  lazy   languages,  where
heap-allocated objects  may include closures  (runtime representations
of unevaluated expressions). Since data  is made live by evaluation of
closures, and  in lazy languages the  place in the program  where this
evaluation  takes  place  cannot be  statically  determined,  laziness
complicates  liveness  analysis  itself.    In  addition,  apart  from
evaluated   data,  we  need to  extend   liveness-based  collection   to
closures i.e. we reclaim the space for closures  which have no future  use.
Since a closure can  escape the scope in which it  was created, it has
to carry the liveness information  of its free variables. Moreover, as
execution  progresses and  possible  future uses  are eliminated,  the
liveness information is updated with a more precise version.  All this
requires significant  modification to  the earlier  garbage collection
scheme for eager languages.

\warning{Experiments  with  a   single  generation  copying  collector
  (Section~\ref{sec:experiments})  confirm  the  expected  performance
  benefits.   Liveness-based  collection  results in  an  increase  in
  garbage reclaimed.   As a  consequence, there is  a decrease  in the
  number of  collections (up to 45\%)  and a decrease (up  to 88\%) in
  the number  of {\em dragged}  cells (cells  that survive a  GC after
  their  last use).  Besides,  there  is a  reduction  in the  overall
  execution time for some programs.  Our experiments show that a mixed
  strategy, where we  do liveness based GC up to  the closure boundary
  and reachability based GC for  the arguments (free variables) of the
  closure,  gives us  similar benefits  as a  fully liveness  based GC
  (where  the  arguments  of  the closure  are  also  collected  using
  liveness) with lesser overhead (Section~\ref{sec:strategies}).}

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\begin{figure}[t!]
  \psset{unit=1mm}
  \begin{pspicture}(90,52)(0,-52)
    %\psframe(90,48)(0,-48)
    \renewcommand{\arraystretch}{1}
    \begin{tabular}{@{}c@{}}
      %\psframebox
          {\sf
	\renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
            \UFL\
            \UNL{0} (\DEFINE\ (\length\  \pl)
	    \UNL{1}  (\SIF~(\NULLQ \ \pl) $0$
            (\PRIM\ 1\ (\length\  (\CDR\  \pl)))))
            \UNL{0}
	    \UNL{0}  (\DEFINE\ (\append\  \lista\ \listb)
	    \UNL{1}  (\SIF~(\NULLQ \ \lista)
	    \listb
	    \UNL{2} \hspace*{-0.1cm}(\CONS\ (\CAR\ \lista) (\append\
            (\CDR\  \lista)
            \listb))))
            \UNL{0}
            \UNL{0} (\LET\ \px\
            $\leftarrow$\ (\CONS\ $5$
            (\CONS\ (\CONS\ $6$ \NIL) \NIL) \IN
	    \UNL{1} (\LET\ \py\   $\leftarrow$\  (\CONS\ $3$ \NIL) \IN
	    \UNL{2}
            (\LET\ \pz\  $\leftarrow$\  (\append\ \px\  \py)\ \IN\
            \UNL{3} (\SIF~(\NULLQ~(\CAR~\pz))~$0$~$\pi$:\
            (\length\ \pz))))))
	  \end{uprogram}
      }}
      \\ \\
      (a) Example program. \\  \\
%      \includegraphics[width=.45\textwidth]{motiv-example}
      \input{mem-graph} \\ \\
      \renewcommand{\arraystretch}{.9}
      \begin{tabular}[t]{p{0.9\columnwidth}}
        (b) Memory graph at $\pi$.  \scalebox{.7}{\TwoCellsAD{a1}{a2}}
        denotes a closure. Thick edges denote live links. Traversal
        stops at edges marked $\times$ during garbage collection for
        a liveness-based collector.
      \end{tabular}
    \end{tabular} 
  \end{pspicture}
%  \vspace*{-3ex}
  \caption{Example Program and its Memory Graph}\label{fig:mot-example}
\end{figure}

Figure~\ref{fig:mot-example} shows an example program and the state of
the heap at  the program point $\pi$, i.e. just  before the evaluation
of $(\length\  \pz)$.  The heap is  represented by a graph  in which a
node either  represents atomic  values ($\NIL$,  integers etc.),  or a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent  variables  or fields.  The figure shows
the  lists \px\  and  \pz\ partially  evaluated.   The evaluation  was
triggered by the test (\NULLQ~(\CAR~\pz)) in the \SIF\ expression.

The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus if a garbage collection  takes place at
$\pi$ with the heap shown in Figure~\ref{fig:mot-example}(b), a
liveness-based collector (LGC) will preserve only the cells referenced
by  $\pz$,  and the live cells  constituting the  closure
referenced by $(\CDR~\pz)$.   In contrast, a reachability-based
collector (RGC) will preserve all cells.

In  this work  we show  that  static analysis  of heap  data can  help
garbage collectors in reclaiming more garbage.  The specific
contributions of this paper are
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first  order
  functional language and  prove its correctness. The correctness
  proof involves setting up a non-standard semantics as a
  specification of liveness and then proving that the analysis is
  correct with respect to the specification.   
\item Our  analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  regarding  whether   to  copy  a  cell   during  garbage  collection
  translates  to the  membership problem  of such  grammars.  In  this
  paper, we  have shown the  undecidability of the  membership problem
  for  the  class  of  grammars  that result  from  our  analysis.  To
  circumvent the  problem, we  over-approximate the  grammars generated
  by the analysis with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  demonstrate the effectiveness of liveness-based garbage collection.
\end{itemize}
To the  best of our
  knowledge, this is the first work on liveness-based garbage
  collection      for      lazy     languages.       All      previous
  attempts~\cite{shaham01heap,   ran.shaham-sas03,   shaham02estimating,
    asati14lgc, karkare06effectiveness} have targeted either
  imperative or  eager functional languages.

@@

\warning{Section~\ref{sec:defs}   introduces   the   syntax   of   the
  programming language  considered and gives a  small-step operational
  semantics for it.   The liveness analysis for this  language and its
  soundness   proof   is  presented   in   Section~\ref{sec:liveness}.
  Section~\ref{sec:computing} describes the formulation of liveness as
  grammars. We  also give a  proof of undecidability of  such grammars
  and    show    how   they    can    be    approximated   by    DFAs.
  Section~\ref{sec:GC-scheme}   discusses  details   of  the   garbage
  collector,  in  particular the  use  of  liveness DFAs  for  garbage
  collection.     We    report    our    experimental    results    in
  Section~\ref{sec:experiments}   along    with   some   observations.
  Section~\ref{sec:relatedwork}  discusses  previous work  related  to
  garbage  collection  and liveness  and  Section~\ref{sec:conclusion}
  discusses possible extensions and concludes the paper.}
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective} where  all actual  parameters to
functions  are  variables.  While  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of  closures through  the $\LET$  construct.  \warning{ For
  now, we assume  $\LET$s  in our
language  are non-recursive;  in the  expression $\LET\,\,  x \leftarrow  s\,\,
\IN\,\, e$, $x$ should not occur in  $s$}. The  restriction of \LET\ to  a single
definition  is for  ease  of  exposition---generalization to  multiple
definitions does not add conceptual difficulties.  We further restrict
each variable in a program to  be distinct, so that no scope shadowing
occurs---this  simplifies  reasoning.

The body  of a function ${\mathit  f}$ is denoted  as $e_{\mathit f}$.
 We assume that each program has a distinguished function
\mainpgm\      with      the      definition     $(\DEFINE\      ({\tt
  \mainpgm})\  e_\mainpgm)$ and  the execution  of the  program starts
with the call to \mainpgm.  We write $\pi\!:\!e$ to associate the label
$\pi$ (not  part of the language  syntax) with the  program point just
before expression $e$.

\begin{figure}[t]\footnotesize
\renewcommand{\arraystretch}{1}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\,\,\, e_\mainpgm
   \hspace{5em} \,\,\,\,\,\,\,\,\,\; \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \ \ \ \ \  \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
  \caption{The syntax of our language}\label{fig:lang-syntax}
\figrule
\normalsize
\end{figure}

\begin{figure*}[t!]
%\comment{

\begin{center}\footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline 
          & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\, \kappa
  \rightsquigarrow \rho',\, S,\, H[\ell := \kappa],\, e$    &  \sc{const}
\\
\hline
          & {$\rho, \,(\rho', \ell, e)\!:\!S,\, H,\, (\CONS~x~y)
\rightsquigarrow
$  $\rho',\, S,\, H[\ell := (\rho(x), \rho(y))],\, e$}     &  \sc{cons} \\
\hline

$H(\rho(x)) \mbox{ is } (v, d)$ & $\rho,\, (\rho', \ell, e)\!:\!S,\, H,\,
(\CAR~x)  \rightsquigarrow \rho',\, S,\, H[\ell := v],\, e$      &
\sc{car-select} \\
\hline
